Step by Step: Using Step Selection Analysis to Simulate Dispersal and Assess Landscape Connectivity

### 2021-03-22 Chapter_1/04_Manuscript/Manuscript.tex
Various updates

\textbf{Running Title:} Simulating Wild Dog Dispersal.

@@

\textbf{Keywords:} dispersal, habitat selection, integrated step selection
function, Kavango-Zambezi Transfrontier Conservation Area, landscape
connectivity, least-cost corridors, Lycaon pictus, permeability surface,
protected areas, wildlife management

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
This is the abstract.
\end{abstract}

@@

\section{Introduction}
An animals movement trajectory can be seen as the result of an interplay between
habitat and movement preferences.

@@

@@

Reliable identification of dispersal corridors will become increasingly
important with the uprise of ever-growing and often transboundary conservation
areas. One such instance is the KAZA-TFCA, a massive conservation area spanning
five countries and over 520'000 km\textsuperscript{2}. The KAZA holds the
potential of re-establishing dispersal routes for many of its protected species,
including the african wild dog \textit{Lycaon pictus}. This species has
experienced severe population declines caused by human induced mortality and
deadly diseases. In result, the species currently marks the KAZA's most
endangered large carnivore and has been a assigned a very high conservation
priority. Importantly, due to their inherent mobility and intrinsic need for
vast undisturbed landscapes, AWDs have been proposed as surrogate species for
landscape connectivity (see recent paper on multispecies connectivity).
Nevertheless, the species has received little attention in the connectivity
literature, mainly due to the difficulty in observing wild dog dispersal. In a
previous paper, we addressed this issue and developed a habitat selection model
based on which we predicted landscape connectivity using least-cost corridor
analysis. We now expand on this knowledge and develop a more detailed movement
model of dispersing wild dogs. We then use this model to simulate thousands of
dispersers moving throughout the KAZA. Based on said simulations, we compute
heatmaps and identify potential dispersal hotspots and compare them to the
dispersal routes identified in (Hofmann ....). We also apply the data to feed a
network-analysis, based on which we determine network-metrics pertinent to
landscape connectivity.

\section{Methods}
\subsection{Study Area}
The study area (centered at -17\degree 13'9''S, 23\degree 56'4''E;
\Cref{StudyArea}a) stretched over 1.3 Mio km\textsuperscript{2} and ecompassed
the entire KAZA-TFCA (\Cref{StudyArea}b). The KAZA-TFCA is the world's largest
transboundary conservation area and comprises parts of Angola, Botswana,
Namibia, Zimbabwe, and Zambia, covering a total area of over 520'000
km\textsuperscript{2}. Its landscape varies regionally and ranges from savanna,
to grassland, and from dry to moist woodland habitats. A dominant
hydrogeographical feature in its center is the Okavango Delta, the earth's
largest inland delta and home to a vast diversity of mammal species. The delta
and its surroundings are considered a stronghold for African wild dogs which may
act as a source for the recolonization of surrounding habitats. The wet season
in the region lasts from November to March, albeit the main floodwaters reach
the Delta between July and August after descending through the Angolian
highlands \citep{McNutt.1996, Wolski.2017}.

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_StudyArea}
    \caption{Study area}
    \label{StudyArea}
  \end{center}
\end{figure}

\subsection{GPS Relocation Data}
Between 2011 and 2020, we collected GPS relocation data on free-ranging wild
dogs inhabiting the Okavango Delta in northern Botswana. We identified potential
dispersers based on criteria reported in \cite{Behr.2020}, immobilized them
according to protocols described in \cite{Osofsky.1996}, and fitted them with
GPS/Satellite radio collars (\textit{Vertex Lite; Vectronic Aerospace GmbH,
Berlin}). Handling and collaring of all individuals was carried out and
supervised by a Botswana-registered wildlife veterinarian. Of all collared
individuals, a total of 16 individuals dispersed, each in a same-sex coalition
(7 female and 9 male coalitions). When we observed dispersal we remotely
increased the GPS fixrate from 24 to 4 hours. Collected relocations were then
regularly transmitted via Iridium satellite system to a base station. This
allowed remote tracking of dispersers even if they left the main study area or
crossed international borders. To effectively distinguish between resident and
dispersersing individuals, we applied the net-squared displacement metric. This
metric measures the Euclidean distance of a collared individual to a reference
point \citep{Borger.2012}. In our case, these reference points resembled the
center of the dispersing coalition's natal home range. Hence, dispersal was
deemed to have started when an individual left its natal home range and ended
when the individual became stationary again. Previous research found no
differences between males and females during dispersal \citep{Woodroffe.2019,
Cozzi.2020}, so we did not distinguish sexes in our analyses. GPS relocations
were then converted to steps, where a step represented the straight-line
distance travelled between two consequtive GPS relocations
\citep{Turchin.1998}.

\subsection{Covariates}
To represent environmental covariates spatially, we prepared a set of raster
layers depicting water-cover (dynamically updated), square rooted distance to
water (dynamically updated), tree-cover, and shrub/grassland-cover. We also
created a proxy for human influence, rendering anthropogenic pressures stemming
from human-density, agricultural sites, and roads. We prepared all layers at a
resolution of 250m by 250m for the extent of the entire KAZA-TFCA. A more
detailed description of the derivation and preparation of each environmental
covariate is given in \cite{Hofmann.2020}. Because wild dogs follow a diurnal
activity pattern, we coded a binary variable indicating whether an observed step
started during phases of main wild dog activity (07:00 a.m. and 07:00 p.m.) or
low wild dog activity (anything else).

\subsection{Movement Model}
We used integrated step selection functions (iSSF) to parametrize a mechanistic
movement model of dispersing wild dogs. In the iSSF framework, observed steps
receive a selection score \(w(x)\) that depends on the covariates \(X\)
experienced along a step, as well as the animals preferences towards these
covariates \(\beta\). In contrast to regular step selection analysis,
\textit{integrated} step selection analysis allows simulatneous inference on
movement and habitat preferences of the studied animal. Moreover, potential
interactions between habitat and movement preferences can be modelled. Thus, the
method produces more accurate selection estimates and allows to use the
resulting models as proper mechanistic movement models from which movement can
be generated. To conduct iSSf analysis, we paired each observed step with 24
random steps. Random steps basically resembled potential alternatives that the
animal could have realized but decided not to. We generated random steps by
sampling random turning angles from a uniform distribution (\(-\pi, +\pi\)) and
step lengths from a gamma distirbution that was fitted to observed steps (scale
= 6308, shape = 0.37). While the number of random steps is inversely
proportional to the sampling error \citep{Avgar.2016}, we found only minor
changes in inference when sampling further random steps. Along each step we
extracted spatial covariates using the \textit{velox} package. We also
calculated step metrics, namely the natural logarithm of the step length
\(log(sl\_\)) and the cosine of the turning angle \(cos(ta\_)\). We scaled all
continuous covariates to a mean of zero and a standard deviation of one. We then
used the r-package \textit{glmmTMB} to fit mixed effects conditional logistic
regression models as proposed by \citep{Muff.2020}. Our movement model was built
around a previously published habitat selection model of dispersing wild dogs.
Because the earlier model was used to predict landscape permeability,
interactions between environmental and spatial covariates were impossible to
model. Our current movement model, in contrast, allowed the inclusion of
additional interactions that describe differences in movement behavior induced
through different environments. Hence, we started with the habitat model and
iteratively increased model complexity by including additional interactions
between movement metrics and environmental covariates. That is, we proposed all
two-way interactions between spatial covariates and movement metrics. For
instance, for the covariate \textit{water} we proposed the interactions
Water:cos(ta\_) and Water:log(sl\_). Despite interactions between environmental
factors and movement metrics, we also proposed the interaction
\(log(sl\_):MainActivity\) to account for the diurnal activity pattern of wild
dogs. A preliminary revealed little differences in turning angles during main
and low activity, hence we did not include the interaction
\(cos(ta\_):MainActivity\). We then ran stepwise model forward selection based
on Akaike's Information Criterion (AIC, \citealp{Burnham.2002}) values and
identified the most parsimonious movement model. Although several models
received an AIC weight above one, we decided to only consider the  ``best''
model for simplicity. In either way, the models with positive weights contained
almost identical covariates and an averaged model would have only given weak
support to additional covariates.

\subsection{Dispersal Simulation}
We used the most parsimonious movement model to simulate dispersing wild dogs
departing from predefined source points. The simulation basically resembled an
inverted iSSF function and was set up as follows. First, we determined a source
point at which a disperser was initiated. To allow calculation of turning
angles, we assumed a random initial orientation of the animal. Second, we
proposed 25 random steps originating at the predefined source point. We
generated random steps by sampling turning angles from a uniform distribution
(\(-\pi, +\pi\)) and step lengths from a gamma distribution fitted to observed
step lengths. To prevent unrealistically large steps, we capped the gamma
distribution 35km, the farthest distance travelled in four hours according to
our data. Third, along each random step we extracted environmental covariates
calculated step metrics. Fourth, we applied the parametrized dispersal model to
predict selection scores \(w(x)\) and to determine the probability of a step
being realized. Fifth, we sampled one of the proposed based on their
probabilities and calculated the animal's new position. We then repeated steps
two to five until 2000 steps were realized. In case a proposed random step left
our study extent, we removed the step from the set of random steps, thereby
forcing the disperser to stay within the boundaries of the study area.

\subsection{Source Points}
Simulations from step selection functions are known to be sensitive to the
location of source points, which is why we followed a twofold approach to sample
release points across the KAZA-TFCA. In a first approach, we used the same 68
source points between which we already computed least-cost paths and least-cost
corridors (see xx). These source points were regularly spaced 100 km apart,
located within protected areas (\(>\) 700 km\textsuperscript{2}). Because of
the regulary spacing, not all protected areas received source points. Hence, we
placed additional source points at the center of such protected areas. At each
of the generated source point we released 1000 dispersers, implying a total of
68'000 simulated dispersers. In a second approach, we used the same 68 source
points to define catchment areas inside protected areas. That is, for each of
the 68 source points we defined its voronoi polygon within the point's protected
area. Within these catchment areas we then randomly sampled 1000 new source
points from which we released another 68'000 dispersers. Overall, we simulated
134'000 dispersers for a total of 268 Mio. steps.

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_SourcePoints}
    \caption{Illustration of source points from which dispersal was simulated.
    Although in reality we simulated 1'000 per source point, we illustrate an
    example assuming 10 dispersers from each source point. (a) Static source
    points similar to the source points reported in ... (b) Source points that
    were randomized within the catchment areas (dark gray, delinated by solid
    black lines).}
    \label{SourcePoints}
  \end{center}
\end{figure}

\subsection{Heatmaps}
Using the simulated trajectories we created heatmaps to examine through which
areas most simulated individuals dispersed. We rasterized all trajectories and
calculated how often each raster-cell was traversed by simulated dispersers. If
the same trajectory crossed a pixel twice, it was only counted once. We achieved
high performance rasterization of spatial lines using the recently developed
R-package \textit{terra} \citep{Hijmans.2020}. To examine if and how ``heat''
changes in response to changes in the location of source points and the number
of simulated steps, we followed a 2 x 6 design and created heatmaps for both
point sampling regines, as well as for 68, 125, 250, 500, 1000, and 2000
dispersal steps. We quantified the similarity of the resulting 12 heatmaps to
the permeability and least-cost corridor maps presented in (Hofmann ...) we used
Bhattacharyya's affinity. Bhattacharyya's affinity ranges from zero (complete
separation) to one (perfect match) and has earlier been proposed to compare the
overlap of utilisation distributions (Fieberg).

@@

\subsection{Network Analysis II: Betweenness}
We coerced all simulated trajectories into a network consisting of vertices
(relocations) and edges (connections between relocations). To do so, we created
raster layers at multiple resolutions and identified each trajectory's
transition matrix on these rasters see figure xx \citep{Bastille.2018}. We then
merged the transition matrices of all trajectories and calculated cumulative
transitions between all raster-cells. This resulted in an edge-list, containing
all observed from-to connections as well as their frequency. Using this
edge-list, we generated a weighted graph using the r-package \textit{igraph}.
Based on this graph we calculated betweenness scores as well as the degree of
each raster cell. Betweenness indicates how often a specific raster-cell lies on
a shortest path between two other raster-cells and is a useful metric to detect
movement corridors. Degree, on the other hand, indicates how many connections a
raster cell has and therefore serves to illustrate the xxx of specific nodes.
When calculating betweenness, we used the transition frequency as weighting
factor. That is, a higher transition frequency contributed to a higher
betweenness score.

@@

\begin{figure}
  \begin{center}
    \begin{subfigure}[c]{0.8\textwidth}
      \includegraphics[width=\textwidth]{99_MovementModel}
      % \subcaption{Subfigure Bild Nr. 1}
    \end{subfigure}
    % \begin{subfigure}[c]{\textwidth}
    %   \includegraphics[width=\textwidth]{99_MovementModel(Interactions)}
    %   % \subcaption{Subfigure Bild Nr. 2}
    % \end{subfigure}
    \caption{Most parsimonious movement model. Whiskers delineate the 95\%
    Confidence-Intervals. Significance codes: * \(p < 0.10\), ** \(p < 0.05\),
    *** \(p < 0.01\).}
    \label{MovementModel}
  \end{center}
\end{figure}

@@

@@

>>>>

\textbf{Running Title:} Release the Dogs! Simulating Wild Dog Dispersal to
Assess Landscape Connectivity

@@

\textbf{Keywords:} dispersal, simulation, integrated step selection,
Kavango-Zambezi Transfrontier Conservation Area, landscape connectivity, Lycaon
pictus

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}

During the past two decades, least-cost analysis and circuit theory have been
the two workhorses in connectivity analyses. This is largely thanks to their
ease of use and intuitive nature which has facilitated the application of the
methods across a broad range of animals. However, recent innovations in movement
ecology have brought forward new methods that are similarly intuitive but
provide several advantages over traditional connectivity modelling techniques.

Here, we propose the simulation of dispersal trajectories as a much more generic
way to identify dispersal barriers and potential movement corridors. Moreover,
we show how such simulations can be used to identify zones of high
human-wildlife conflicht, as well as to pinpoint potential "dispersal traps"
where dispersing individuals get trapped due to unavailability of suitable
alternatives. Finally, we discuss the benefits and pitfalls of dispersal
simualtions.
\end{abstract}

@@

\section{Introduction}
% Dispersal & Connectivity
Dispersal of individuals is an important process governing the dynamics wild
animal populations that are distributed in space \citep{Clobert.2012}. It is
defined as the movement of individuals from their natal location to the site of
first reproduction \cite{Howard.1960} and allows to avoid inbreeding
\citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012}, to rescue small and
unviable populations \citep{Brown.1977}, and to promote the colonization of
unoccupied habitats \citep{Hanski.1998, MacArthur.2001}. Information on movement
behavior during dispersal and knowledge about the factors that limit dispersal
is therefore critical for a comprehensive understanding of landscape
connectivity and population viability \citep{Baguette.2013, Vasudev.2015}.

% Technological Advancements
Thanks to novel technologies developed over the past decades, particularly of
GPS/Satellite radio-collars, the study of dispersal has accelerated
\citep{Jonsson.2016, Williams.2019}. Additionally, the advent of publicly
accessible satellite imagery and sophisticated remote sensing techniques to
represent the physical landscape through which individuals disperse has pushed
the field towards a  ``golden age of animal tracking'' \citep{Kays.2015}.
Concurrently, the increased availability of large amounts of empirical data and
an increased computational power have led to the development of numerous
analytical methods to study movement behavior during dispersal
\citep{Boyce.2002, Fortin.2005, Cushman.2010, Zeller.2012, Elliot.2014,
Diniz.2019}.

% Measuring Connectivity
Resource selection functions \citep{Boyce.2002} and derived methods such as step
selection functions \citep{Fortin.2005} and path selection functions
\citep{Cushman.2010} have proven particularly useful for studying animal
movement and landscape connectivity. These methods allow estimating habitat
preferences of the focal species by comparing covariates at locations visited by
the animal to the scame covariates at locations available to, but not visited by
the animal. The so estimated preferences can then be used to predict a
permeability surface, indicating the expected ease at which an animal can
traverse a given area \citep{Zeller.2012}. Utimately, the permeability surface
serves as foundation to a connectivity model and to reveal movement corridors.
In this regard, the literature has long relied on least-cost methods
\citep{Adriaensen.2003} and circuit theory \citep{McRae.2006, McRae.2008}.
Unfortunately, both of these approaches make several limiting assumptions that
are hardly ever met in reality, particularly by dispersing animals. For
instance, to calculate a least-cost path one has to pre-define start- and
end-points. This implicitly assumes that animals have a preconceived end-point
in mind and choose a cost-minimizing route accordingly. To calculate a
least-costly route, one also needs to assume that animals have an infinite
perceptual range and that a route can always be realized, regardless of the
imposed costs. In contrast to least-cost methods, circuit theory assumes a
perceptual range of a single pixel, which is equally unrealistic and rarely
captures the scale of selection (cite someone). Moreover, directional biases
cannot be captured by circuit theory as a random walk is implicitly assumed. In
reality, however, dispersers often depict directional persistence and move in an
almost straight line to cover much ground very quickly. A mix of the two methods
can be seein in least-cost corridor approaches or randomized shortest paths have
been proposed to mitigate these issues by allowing animals to deviate from the
``optimal'' least-costly route \citep{Pinto.2009, Panzacchi.2016,
VanMoorter.2021}, they still fail to render a more realistic perceptual range of
the animal. In addition, neither of the two methods render the temporal
dimension of movement. Thusfar, only few have explored the usefulness of
individual based movement models when estimating landscape connectivity
\citep{Kanangaraj.2013, Hauenstein.2019, Zeller.2020}.

Nevertheless, collecting data on dispersers remains logistically challenging due
to dispersers' inherent mobility and due to the fact that dispersers only make
up a small fraction of the entire population, thereby making it difficult for
scientists to target apporpriate individuals for their study.

An animals movement trajectory can be seen as the result of an interplay between
habitat and movement preferences.

@@

However, these models can not account for a correlation between turning angles
and step lengths, unless the two are sampled jointly from a copula distribution
(Hodel et al.).

A relatively recent addition to the step selection family has been integrated
step selection analysis.... Thanks to recently published papers and packages,
the suite of step selection functions has become accessible to most researchers
working with relocation data. Despite this, the use of such methods rarely
extends beyond assessing relative habitat preferences. However, once
parametrized, integrated step selection models could potentially serve as
powerful framework to simulate animal movement and assess landscape
connectivity. Here, we exemplify such analyses and show how ... coud be applied
to investigate ...

In contrast to Signer et et al, our focus lied on the transient utilisation of
the landscape.

@@

Reliable identification of dispersal corridors will become increasingly
important with the uprise of ever-growing and often transboundary conservation
areas. One such instance is the KAZA-TFCA, a massive conservation area spanning
five countries and over 520'000 km\textsuperscript{2}. The KAZA holds the
potential of re-establishing dispersal routes for many of its protected species,
including the african wild dog \textit{Lycaon pictus}.  Persecution by humans,
habitat loss, and reduced connectivity are major causes of the decline of the
species \citep{Woodroffe.2012}. In result, the species currently marks the
KAZA's most endangered large carnivore and has been a assigned a very high
conservation priority. Importantly, due to their inherent mobility and intrinsic
need for vast undisturbed landscapes, AWDs have been proposed as surrogate
species for landscape connectivity (see recent paper on multispecies
connectivity). Nevertheless, the species has received little attention in the
connectivity literature, mainly due to the difficulty in observing wild dog
dispersal. In a previous paper, we addressed this issue and developed a habitat
selection model based on which we predicted landscape connectivity using
least-cost corridor analysis. We now expand on this knowledge and develop a more
detailed movement model of dispersing wild dogs. We then use this model to
simulate thousands of dispersers moving throughout the KAZA. Based on said
simulations, we compute heatmaps and identify potential dispersal hotspots and
compare them to the dispersal routes identified in (Hofmann ....). We also apply
the data to feed a network-analysis, based on which we determine network-metrics
pertinent to landscape connectivity.

\section{Methods}
\subsection{Study Area}
The study area (centered at -17\degree 13'9''S, 23\degree 56'4''E;
\Cref{StudyArea}a) stretched over 1.3 Mio km\textsuperscript{2} and ecompassed
the entire KAZA-TFCA (\Cref{StudyArea}b). The KAZA-TFCA is the world's largest
transboundary conservation area and comprises parts of Angola, Botswana,
Namibia, Zimbabwe, and Zambia, covering a total area of over 520'000
km\textsuperscript{2}. Its landscape varies regionally and ranges from savanna,
to grassland, and from dry to moist woodland habitats. A dominant
hydrogeographical feature in its center is the Okavango Delta, the earth's
largest inland delta. The delta and its surroundings are considered a stronghold
for African wild dogs and may act as a source for the recolonization of
surrounding habitats. The wet season within our study area lasts from November
to March and is out of phase with the main flooding of the Okavango Delta which
peaks between July and August \citep{McNutt.1996, Wolski.2017}.

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_StudyArea.png}
    \caption{Study area}
    \label{StudyArea}
  \end{center}
\end{figure}

\subsection{GPS Relocation Data}
Between 2011 and 2020, we collected GPS relocation data on dispersing wild dogs
from a free-ranging wild dog population inhabiting the Okavango Delta in
northern Botswana. We identified potential dispersers based on age, number of
same‐sex siblings, pack size, and presence of unrelated individuals of the
opposite sex in their pack \citep{McNutt.1996, Behr.2020}. We immobilized
selected individuals using a cocktail of Ketamine/Xylazine/Atropine
\citep{Osofsky.1996, Cozzi.2020} which we injected with a dart, fired from a
CO\textsubscript{2}-pressurized gun (\textit{DAN-Inject, Denmark}). The
corresponding protocols have been approved by the Ministry of Environment,
Natural Resources Conservation and Tourism of Botswana (permit EWT 8/36/4
XXXVI). After immobilization, individuals were fitted with GPS/Satellite radio
collars (\textit{Vertex Lite; Vectronic Aerospace GmbH, Berlin}) that included
an automated drop-off mechanism. Handling and collaring of all individuals was
carried out and supervised by a Botswana-registered wildlife veterinarian. All
of the immobilized individuals quickly rejoined their pack after the procedure.
From all collared individuals, 16 individuals eventually dispersed in separate
same-sex coalitions and their trajectories were successfully recorded (7 female
and 9 male coalitions).

During dispersal, collars were programmed to record a GPS fix every 4 hours.
Recorded relocations were regularly transmitted over the Iridium satellite
system, which allowed remote tracking of individuals, even if they left the main
study area and crossed international borders. To effectively distinguish between
periods of residency and dispersal, we applied the net-squared displacement
metric to the observed movement data. This metric measures the squared Euclidean
distance of a collared individual to a reference point \citep{Borger.2012},
which we set at the center of each disperser's natal home range. Hence,
dispersal was deemed to have started when an individual left its natal home
range and ended when the individual became stationary again. For the purpose of
this study, we discarded andy data that was collected during residency. Because
previous research could not find any differences between males and females
during dispersal \citep{Woodroffe.2019, Cozzi.2020}, we did not distinguish
sexes for the sake of our analyses. Ultimately, we converted the collected GPS
relocations to steps, where a step represented the straight-line distance
travelled between two consequtive GPS relocations \citep{Turchin.1998}.

\subsection{Covariates}
To represent habitat covariates, we prepared a set of spatial raster layers
depicting water-cover (dynamically updated), distance to water (dynamically
updated), tree-cover, and shrub/grassland-cover. To render a decreasing impact
with increasing distance, we square rooted the distance to water covariate
layer. We also created a proxy for human influence, rendering anthropogenic
pressures stemming from human-density, agricultural sites, and roads. We
rendered all layers at a resolution of 250m by 250m for the entire extent of the
KAZA-TFCA. Further details on the derivation and preparation of each
environmental covariate is given in \cite{Hofmann.2020}.

Besides habitat covariates, we also computed movement metrics that we used as
movement covariates in our models. These covariates included the step length
(\(sl\)), its natural logarithm (\(log(sl)\)), as well as the cosine of the
relative turning angle (\(cos(ta)\)). Because wild dogs follow a diurnal
activity pattern, we furthermore coded a binary variable indicating whether an
observed step was realized during main activity (xx to xx) or low activity (xx
to xx).

\subsection{Movement Model}
We used integrated step selection functions (iSSF) to parametrize a mechanistic
movement model of dispersing wild dogs. In this framework, observed steps are
contrasted to random steps that the animal could have realized, but decided not
to. To compare steps, it is assumed that animals assign a relative selection
score \(w(x)\) to each step. The score depends on the covariates \(X\)
experienced along a step, as well as the animals preferences towards these
covariates \(\beta\). The probability of a step being realized is then
contingent on the step's selection score, as well as the selection scores of all
other step in the same stratum. In contrast to regular step selection analysis,
\textit{integrated} step selection analysis allows simulatneous inference on
movement and habitat preferences of the studied animal. Moreover, potential
interactions between habitat and movement preferences can be included to
investigate how movement behavior depends on habitat characteristics. In result,
the method produces less biased selection estimates and allows to apply the
resulting model as a fully mechanistic movement model based on which movement
can be simulated. To conduct iSSF analysis, we paired each observed step with 24
random steps. Random steps basically resembled potential alternatives that the
animal could have realized but decided not to. We generated random steps by
sampling random turning angles from a uniform distribution (\(-\pi, +\pi\)) and
step lengths from a gamma distirbution that was fitted to observed steps (scale
= 6'308, shape = 0.37). While the number of random steps is inversely
proportional to the sampling error \citep{Avgar.2016}, we found only minor
differences in model performance when sampling additional random steps and
deemed 24 random steps sufficient. Along each step, we extracted spatial
covariates using the \textit{velox} package and calculated the movement metrics
\(sl\), \(log(sl\)), and \(cos(ta)\). To facilitate model convergence, we
standardized all continuous covariates to a mean of zero and a standard
deviation of one. We then used the r-package \textit{glmmTMB} to fit mixed
effects conditional logistic regression models as proposed by \citep{Muff.2020}.

Our movement model was and extension of the habitat model presented in Hofmann
et al 2021. We will refer to it as the base model. Because the base model was
used to predict landscape permeability, no interactions between the habitat and
movement metrics were modeled. Thus, we expanded the base model and by proposing
additional interactions between movement metrics and environmental covariates.
For this, we started with the base model and iteratively increased model
complexity by adding all proposed two-way interactions between habitat
covariates and movement covariates. For instance, for the covariate
\textit{water} we proposed the interactions \(Water:log(sl)\),
\(Water:log(sl)\), and \(Water:cos(ta)\). Furthermore, we proposed the
interactions \(sl:cos(ta)\) and \(log(sl):cos(ta)\) to account for a correlation
between turning angles and step lengths, as well as the interactions
\(sl:MainActivity\) and \(log(sl):MainActivity\) to account for the fact that
step lengths diffuer due to the diurnal activity pattern. We then ran stepwise
model forward selection based on Akaike's Information Criterion (AIC,
\citealp{Burnham.2002}) values and identified the most parsimonious movement
model. Although several models received an AIC weight above one, we decided to
only consider the  ``best'' model for simplicity. In either way, the models with
positive weights contained almost identical covariates and an averaged model
would have only given weak support to additional covariates.

\subsection{Source Points}
We randomly placed source points within protected areas larger (\(>\) 700
km\textsuperscript{2}), which conforms to the average home range requirement of
resident wild dogs \citep{Pomilia.2015}. By distributing source points randomly,
the number of source points per km\textsuperscript{2} within protected areas was
approximately equal across our study area. In total, we distributed 50'000
source points across all protected areas, each representing the starting point
of a dispersal trajectory. To render potential immigrants into the study system,
we randomly placed 10'000 additional source points inside a xx km buffer around
the study area. Hence, a total of 60'000 source points were generated, thereby
resulting in 60'000 simulated dispersers. It is worth pointing out that our
choice of  placing source points completely random is arbitrary, as one as well
adjust the sampling frequency based on the density of dispersers in the
respective source area.

% Simulations from step selection functions are known to be sensitive to the
% location of source points, which is why we followed a twofold approach to sample
% release points across the KAZA-TFCA. In a first approach, we used the same 68
% source points between which we already computed least-cost paths and least-cost
% corridors (see xx). These source points were regularly spaced 100 km apart,
% located within protected areas (\(>\) 700 km\textsuperscript{2}). Because of
% the regulary spacing, not all protected areas received source points. Hence, we
% placed additional source points at the center of such protected areas. At each
% of the generated source point we released 1000 dispersers, implying a total of
% 68'000 simulated dispersers. In a second approach, we used the same 68 source
% points to define catchment areas inside protected areas. That is, for each of
% the 68 source points we defined its voronoi polygon within the point's protected
% area. Within these catchment areas we then randomly sampled 1000 new source
% points from which we released another 68'000 dispersers. Overall, we simulated
% 134'000 dispersers for a total of 268 Mio. steps.

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_SourcePoints.png}
    \caption{Illustration of source points from which dispersal was simulated.
    Although in reality we simulated 1'000 per source point, we illustrate an
    example assuming 10 dispersers from each source point. (a) Static source
    points similar to the source points reported in ... (b) Source points that
    were randomized within the catchment areas (dark gray, delinated by solid
    black lines).}
    \label{SourcePoints}
  \end{center}
\end{figure}

\subsection{Dispersal Simulation}
We used the most parsimonious movement model to simulate dispersing wild dogs
departing from the above described source points. The simulation resembled an
inverted iSSF and was set up as follows. First, we determined a source point at
which a disperser was initiated. To allow calculation of turning angles, we
assumed a random orientation of the animal. Second, we proposed a set of 25
random steps leaving from the source point. We generated random steps by
sampling turning angles from a uniform distribution (\(-\pi, +\pi\)) and step
lengths from our fitted gamma distribution. To prevent unrealistically large
steps, we capped sampled step lengths at 35km, which is the farthest distance
travelled by our dispersers within 4 hours. Third, along each random step we
extracted habitat covariates and calculated movement covariates. Fourth, we
applied the parametrized movement model to predict the selection scores \(w(x)\)
for all steps. Fifth, we translated selection scores into probabilitities by
applying equation xy. Sixth, we sampled one of the proposed based on assigned
probabilities and calculated the animal's new position. We then repeated steps
two to six until a total of 2000 steps were realized.

To mitigate the influence of edge effects, we added a buffer zone around each
covariate layer. The values within the buffer zone where resampled from the true
data. As such, simulated dispersers were allowed to leave and reenter the study
area through this buffer zone. In case one of the proposed random steps left the
outer boundary of the buffer, we replaced the step with an alternative step that
lied fully within the buffer zone. Thus, we force the disperser to bounce of the
borders of the study area.

\subsection{Heatmaps}
Using the simulated trajectories, we created heatmaps that indicate the
frequency at which each raster-cell in the study area was visited. If the same
trajectory crossed a pixel twice, it was only counted once, i.e. we did not
consider revisits. We achieved high performance rasterization of all simulated
trajectories using the recently developed R-package \textit{terra}
\citep{Hijmans.2020}.

To examine if and how ``heat'' changes in response to changes in the location of
source points and the number of simulated steps, we followed a 2 x 6 design and
created heatmaps for both point sampling regines, as well as for 68, 125, 250,
500, 1000, and 2000 dispersal steps. We quantified the similarity of the
resulting 12 heatmaps to the permeability and least-cost corridor maps presented
in (Hofmann ...) we used Bhattacharyya's affinity. Bhattacharyya's affinity
ranges from zero (complete separation) to one (perfect match) and has earlier
been proposed to compare the overlap of utilisation distributions (Fieberg).

\subsection{Betweenness}

@@

\subsection{Network Analysis II: Betweenness}
We coerced all simulated trajectories into a network consisting of vertices
(relocations) and edges (connections between relocations). To do so, we created
raster layers at multiple resolutions and identified each trajectory's
transition matrix on these rasters see figure xx \citep{BastilleRousseau.2018}.
We then merged the transition matrices of all trajectories and calculated
cumulative transitions between all raster-cells. This resulted in an edge-list,
containing all observed from-to connections as well as their frequency. Using
this edge-list, we generated a weighted graph using the r-package
\textit{igraph}. Based on this graph we calculated betweenness scores as well as
the degree of each raster cell. Betweenness indicates how often a specific
raster-cell lies on a shortest path between two other raster-cells and is a
useful metric to detect movement corridors. Degree, on the other hand, indicates
how many connections a raster cell has and therefore serves to illustrate the
xxx of specific nodes. When calculating betweenness, we used the transition
frequency as weighting factor. That is, a higher transition frequency
contributed to a higher betweenness score.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{99_MovementModel}
    \caption{(a) Most parsimonious movement model for dispersing wild dogs. The
    model includes a habitat kernel, a movement kernel, as well as their
    interactions. The orange and blue line segments delineate the 90\%, 95\%,
    and 99\% Confidence-Intervals around the respective \(\beta\) coefficients.
    Significance codes: * \(p < 0.10\), ** \(p < 0.05\), *** \(p < 0.01\). (b)
    Results from the k-fold cross validation procedure. The upper plot shows
    rank frequencies of predicted realized scroes according to model predictions
    with known preferences, whereas the lower plot shows rank frequencies when
    assuming random preferences. The blue ribbon shows the prediction interval
    around a loess smoothing regression that we fitted to ease the
    interpretation of the plots.}
    \label{MovementModel}
  \end{center}
\end{figure}

@@

In this regard, our approach is rather similar to dispersal kernels, yet it
enables to render directional biases, which are currently difficult to implement
using such methods.

We completely randomized the location of source points within protected areas.
However, in some cases prior knowledge about the density of potential dispersers
is available and can be used to adjust the number of simulated individuals
accordingly. Alternatively, instead of tweaking the number of simulated
individuals, one could assign a weight to each trajectory that depends on the
density of potential dispersers in the source areas. As such, trajectories from
areas with high density would enter the heatmap with above average weigth.

The parametrized movement model could also be manipulated to investigate how
different habitat preferences influence landscape connectivity and to test the
sensitivity of results with respect to the exact preferences of individuals.

One of the major benefits of individual-based simulations is the ability to make
the temporal dimension of movement explicit. This allows to investigate how
connectivity depends on the dispersal duration, something that is not possible
with traditional least-cost or circuit theory methods.

Optimally, one should simulate additional dispersers until the amount of newly
gained information lies beyond a certain threshold, i.e. until some sort of
convergence is achieved. However, due to the myriad of outcomes in movement
behavior, this would often imply countless repetitions for each source points at
each location where dispersal is possibly initiated. This is computationally
infeasible.

@@

Comparable simulations could also serve to predict the likely whereabouts of
dispersers as they move. This information can then be utilized as an  ``early
warning'' system, such that potential encounters can be gauged ad hoc.
